```{r}
rmarkdown::pandoc_available()
rmarkdown::pandoc_version()
```


# install packages needed for this notebook


```{r}

library(leaflet)
library(readxl)
library(dplyr)
library(rnaturalearth)
library(ggplot2)
library(sf)
library(giscoR)

```

# Load data

```{r}
data_centers <- read_excel("data/spain_datacenters_v2.xlsx")
```
# Create Spain bounderies

``` {r}

spain <- ne_countries(scale = "medium", country = "Spain", returnclass = "sf")

provinces_raw <- gisco_get_nuts(resolution = "03", nuts_level = 3, year = "2024") %>%
  dplyr::filter(CNTR_CODE == "ES", LEVL_CODE == 3)

provinces_raw %>%
  distinct(NUTS_ID, NAME_LATN) %>%
  arrange(NUTS_ID)




```

# Create map with provinces, exclude the Canary Islands for better visualization of mainland data centers

``` {r}
spain_provinces_map <- ggplot() +
  #Layer 0: spain map
  geom_sf(data = spain, fill = "grey95", color = "grey60") +
  # add boundaries for provinces (excluding canary islands)
  geom_sf(data = provinces_raw, fill = NA, color = "grey80", size = 0.3) +
  theme_minimal(base_size = 11) +
  coord_sf(xlim = c(-10, 5), ylim = c(36, 44)) +
  # add title and theme
  labs(title = "Spain NUTS 3 Provinces",
    subtitle = paste0(n_distinct(provinces_raw$NAME_LATN), " Spanish Provinces (Canary Islands not pictured)"))+
  theme_minimal(base_size = 11) +
  theme(legend.position = "right",
    plot.title = element_text(face = "bold"))
spain_provinces_map
```

``` {r}
ggsave("outputs/spain_provinces_map.jpg", width = 10)
```
# Get datacenter coordinates from data
```{r}
# remove centers where we couldn't get lat/long
data_centers <- data_centers %>%
  filter(!is.na(Latitude) & !is.na(Longitude))

# Radius based on sq meters

data_centers <- data_centers %>%
  mutate(
    size_cat = case_when(
      is.na(`Size (m²)`)              ~ "Unknown",
      `Size (m²)` >= 100000           ~ ">100,000 m²",
      `Size (m²)` >= 10000            ~ "10,000–100,000 m²",
      `Size (m²)` >= 1000             ~ "1,000–10,000 m²",
      TRUE                            ~ "<1,000 m²"
    ),
    size_cat = factor(size_cat, levels = c(
      ">100,000 m²", "10,000–100,000 m²", "1,000–10,000 m²", "<1,000 m²", "Unknown"
    ))
  )
```

 
 # Create the map

``` {r}
data_center_map <- ggplot() +
  #Layer 0: spain map
  geom_sf(data = spain, fill = "grey95", color = "grey60") +
  
  #Layer 1: datacenters with no size info
  geom_point(
    data = data_centers %>% filter(size_cat == "Unknown"),
    aes(x = Longitude, y = Latitude, color = size_cat),
    size = 1.3, alpha = 0.5) +
  # Layer 2:data centers for which we have size data
  geom_point(
    data = data_centers %>% filter(size_cat != "Unknown"),
    aes(x = Longitude, y = Latitude, color = size_cat),
    size = 1.3, alpha = 0.85) +
  scale_color_manual(
    values = c(
      ">100,000 m²"       = "#b10026",
      "10,000–100,000 m²" = "#fd8d3c",
      "1,000–10,000 m²"   = "#41b6c4",
      "<1,000 m²"         = "#225ea8",
      "Unknown"           = "black"),
    name = "Facility Size"
  ) +
  coord_sf(xlim = c(-10, 5), ylim = c(36, 44)) +
  labs(title = "Data Centers in Spain",
    subtitle = paste0(nrow(data_centers), " facilities across ", 
    n_distinct(data_centers$City), " cities")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right",
    plot.title = element_text(face = "bold"))
data_center_map
```

# Save image
```{r}
ggsave("outputs/data_center_map.jpg", width = 10)
```
 
# Zoom in on madrid (accounts for the majority of data centers)
```{r}

# ── NUTS3 boundary for Comunidad de Madrid (ES300) ──
nuts3 <- gisco_get_nuts(resolution = "03", nuts_level = 3, country = "ES", year = "2024")
madrid_nuts3 <- nuts3 %>% filter(NUTS_ID == "ES300")
```


# filter data centers within the Madrid limits

Pulling the 52 provinces across Spain. Provinces are the finest spatial unit with consistent economic and environmental statistics, 
making them the most approapriate unit for this analysis.

```{r}
madrid_limits <- st_bbox(madrid_nuts3)
# Layer 0: zoom on nuts3 comunidad de Madrid
dc_madrid <- data_centers %>%
  filter(
    Longitude >= madrid_limits["xmin"] - 0.2, Longitude <= madrid_limits["xmax"] + 0.2,
    Latitude  >= madrid_limits["ymin"] - 0.2, Latitude  <= madrid_limits["ymax"] + 0.2)

# Layer 1: Plot madrid focus, no size
madrid_dc_map <- ggplot() +
  geom_sf(data = madrid_nuts3, fill = "grey95", color = "grey50") +
  geom_point(data = dc_madrid %>% filter(size_cat == "Unknown"),
    aes(x = Longitude, y = Latitude, color = size_cat),
    size = 2, alpha = 0.5) +
  #Layer 2: plot data centers with size data
  geom_point(
    data = dc_madrid %>% filter(size_cat != "Unknown"),
    aes(x = Longitude, y = Latitude, color = size_cat),
    size = 2, alpha = 0.85) +
  scale_color_manual(
    values = c(
      ">100,000 m²"       = "#b10026",
      "10,000–100,000 m²" = "#fd8d3c",
      "1,000–10,000 m²"   = "#41b6c4",
      "<1,000 m²"         = "#225ea8",
      "Unknown"           = "black"),
    name = "Facility Size") +
  labs(title = "Data Centers in Comunidad de Madrid",
    subtitle = paste0(nrow(dc_madrid), " facilities")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right",plot.title = element_text(face = "bold"))

madrid_dc_map
```
 
 
# Save image
```{r}
ggsave("outputs/madrid_dc_map.jpg", width = 10)
```